---
title: 'Mixed effect Model Repeat Measurement (MMRM)'
author: "Yilong Zhang"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
  flexdashboard::flex_dashboard:
    vertical_layout: fill
  html_notebook:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
---

# Summary

MMRM is commonly used in clinical trials with longitudinal outcome. This page is to illustrate how to fit Mixed effect Model Repeat Measurement (MMRM) in R. 

### Reference

* Siddiqui, O., Hung, H. J., & O'Neill, R. (2009). MMRM vs. LOCF: a comprehensive comparison based on simulation study and 25 NDA datasets. Journal of biopharmaceutical statistics, 19(2), 227-246.
* Mallinckrodt, C. H., Clark, W. S., & David, S. R. (2001). Accounting for dropout bias using mixed-effects models. Journal of biopharmaceutical statistics, 11(1-2), 9-21.
* https://www.slideshare.net/Medresearch/recommendations-for-the-primary-analysis-of-continuous-endpoints
* http://www.biostat.jhsph.edu/~iruczins/teaching/jf/ch5.pdf
* https://stat.ethz.ch/pipermail/r-sig-mixed-models/2015q1/023168.html
* http://www.ics.uci.edu/~staceyah/112-203/R/Stats112-203-RFunctions.R


### Load Function
```{r}
library(mvtnorm)
library(nlme)
library(dplyr)
library(reshape2)
library(lsmeans)
```

```{r}
VarCov_gls <- function(fit){
  vars <- coef(fit$modelStruct$varStruct, uncons = FALSE, allCoef = TRUE)^2 * fit$sigma^2
  r = coef(fit$modelStruct$corStruct, uncons = FALSE, allCoef = TRUE)
  cors = matrix(NA, ncol = length(vars), nrow = length(vars))
	cors[lower.tri(cors)] = r
	cors[upper.tri(cors)] = t(cors)[upper.tri(t(cors))]
	diag(cors) = rep(1, length(vars))
  
  covs <- diag(sqrt(vars) ) %*% cors %*% diag(sqrt(vars))
  rownames(covs) <- names(vars)
  colnames(covs) <- names(vars)
  covs
}
```

### Set up
```{r}
n   <- 100

trt <- c(0, 1.3, 2.3, 3.2, 4)
plc <- c(0, 1.0, 1.8, 2.5, 3)
sd  <- c(2.0, 1.8, 2.0, 2.1, 2.2)

corr <- cor   <- matrix(
  c(1, 0.6, 0.3, 0.2, 0.1, 
    0.6, 1, 0.7, 0.5, 0.2, 
    0.3, 0.7, 1, 0.6, 0.4,
    0.2, 0.5, 0.6, 1, 0.5,
    0.1, 0.2, 0.4, 0.5, 1), 5,5)
cov <- diag(sd) %*% cor %*% diag(sd)
```

### Data Simulation
```{r}
set.seed(123)
y_trt <- rmvnorm(n, mean = trt, sigma = cov)
y_plc <- rmvnorm(n, mean = plc, sigma = cov)
Y <- rbind(y_trt, y_plc)
grp <- rep(1:0, each = n)
db  <- data.frame(id = 1:nrow(Y), Y, grp)
names(db) <- c("id","basval","y1","y2","y3","y4","trt")
db_chg <- db %>% mutate(y1 = y1 - basval,
                        y2 = y2 - basval,
                        y3 = y3 - basval,
                        y4 = y4 - basval)
db_chg_long <- melt(db_chg, id.vars = c("id","basval", "trt"), variable.name = "time")
write.csv(db_chg_long, "db_chg_long.csv")
```


## Full Data Analysis

### Fit ANCOVA for linear model
```{r}
summary(lm( y4 ~ basval + trt, data = db_chg))
```

### MMRM

* ANCOVA and MMRM provide same results when data is completed
* The result for MMRM is equivalent to the SAS code in both completed data and missing data.
* For complete data, the power to compare end timepoint treatment difference should be similar for MMRM and ANCOVA

```
%macro import_csv(data,file);
PROC IMPORT OUT= &data 
            DATAFILE= "&file"
            DBMS=CSV REPLACE;
     GETNAMES=YES;
     DATAROW=2; 
RUN;
%mend import_csv;

%import_csv(tmp, &path\db_chg_long.csv);

proc mixed data=tmp ;
   class trt time id;
   model value= time basval*time trt*time / cl;
   repeated time/type=un subject=id;
   lsmeans time trt*time/cl pdiff;
run;
```

Model Fitting Summary 
```{r}
fit <- nlme:::gls(value ~ relevel(time, ref = "y4") + 
                          basval:relevel(time, ref = "y4") + 
                          factor(trt, levels = c(1,0)): relevel(time, ref = "y4"), 
                   correlation = corSymm( form = ~ 1 | id), 
                   weights = varIdent(form = ~ 1 | time),
                   data = db_chg_long)
fit
```

LS means results
```{r}
## Note that the degree of freedom is wrong in GLS. 
lsmeans(fit, list(pairwise ~ time*trt), df=nrow(db) - 1, adjust = NULL )
```

Variance Covariance Matrix
```{r}
VarCov_gls(fit)
```


## Missing Data Analysis

### Assign Missing Value
```{r}
db_chg_missing <- db
db_chg_missing[c(1:10, n + 1:10), c(3:5)] <- NA
db_chg_missing_long <- melt(db_chg_missing, id.vars = c("id","basval", "trt"), variable.name = "time")
db_chg_missing_long <- na.omit(db_chg_missing_long)
write.csv(db_chg_missing_long, "db_chg_missing_long.csv")
```

Model Fitting Summary 
```{r}
fit <- nlme:::gls(value ~ relevel(time, ref = "y4") + 
                          basval:relevel(time, ref = "y4") + 
                          factor(trt, levels = c(1,0)): relevel(time, ref = "y4"), 
                   correlation = corSymm( form = ~ 1 | id), 
                   weights = varIdent(form = ~ 1 | time),
                   data = db_chg_missing_long)
fit
```

LS means results
```{r}
## Note that the degree of freedom is wrong in GLS. 
lsmeans(fit, list(pairwise ~ time*trt), df=nrow(db) - 1, adjust = NULL )
```

Variance Covariance Matrix
```{r}
VarCov_gls(fit)[c(2:4,1), c(2:4,1)]
```


